---
title: "Bayesian Model for March Madness Score Differential Prediction"
author: "Eftyghia Kourtelidis"
date: "2026-04-04"
output: pdf_document
---

## Bayesian Linear Regression

This notebook predicts tournament point differential using a Bayesian linear 
regression model fit with brms. The model uses pairwise difference features 
between Team 1 and Team 2 derived from regular season statistics, with 
team-level random effects to account for residual team-specific variability.

### Features

For each statistic, features are computed as Team 1 minus Team 2, so each row 
is a vector encoding relative strength between the two teams for that matchup:

* SeedDiff: difference in tournament seed number
* WinPctDiff: difference in regular season win percentage
* ScoreDiffDiff: difference in average regular season point differential
* OffDiff: difference in average points scored (offensive strength)
* DefDiff: difference in average points allowed (defensive strength)
* SeedDiff:WinPctDiff: interaction between seed and win percentage differences

### Preprocessing

* Train/test split by chronological year (2008–2022 train, 2023–2025 test) to 
match the time series split.
* Team assignments (Team 1 vs. Team 2) are randomized to remove any systematic 
bias from winners always appearing on one side in the raw data.
* All features are standardized using training data only, then the same 
transform is applied to test years.

### Model

A Bayesian linear regression model with Gaussian likelihood:

$$y_i \mid w, \sigma, x_i \sim \mathcal{N}(\mu_i = w^T x_i + u_{T1} + u_{T2}, \sigma^2)$$

where $u_{T1}, u_{T2}$ are team-level random intercepts that account for 
residual team-specific variability not captured by the observed features.


### Priors

$$w_j \sim \mathcal{N}(0, 1), \quad \alpha \sim \mathcal{N}(0, 5), \quad \sigma \sim \text{Exponential}(1), \quad u_j \sim \mathcal{N}(0, \tau^2)$$

```{r}
# load libraries
library(brms)
library(tidyverse)
```

``` {r}
# load data
teams <- read.csv("Dataset/MTeams.csv")
regular <- read.csv("Dataset/MRegularSeasonCompactResults.csv")
tourney <- read.csv("Dataset/MNCAATourneyCompactResults.csv")
seeds <- read.csv("Dataset/MNCAATourneySeeds.csv")

# ensure IDs are numeric
regular$WTeamID <- as.numeric(regular$WTeamID)
regular$LTeamID <- as.numeric(regular$LTeamID)

seeds$TeamID <- as.numeric(seeds$TeamID)

head(teams)
head(regular)
head(tourney)
head(seeds)
```

```{r}
# build team statistics
# score differential
regular <- regular %>%
  mutate(ScoreDiff = WScore - LScore)

team_stats <- bind_rows(

  # wins
  regular %>%
    group_by(Season, TeamID = WTeamID) %>%
    summarise(
      Games = n(),
      Wins = n(),
      PointsFor = mean(WScore),
      PointsAgainst = mean(LScore),
      ScoreDiff = mean(ScoreDiff),
      .groups = "drop"
    ),

  # losses
  regular %>%
    group_by(Season, TeamID = LTeamID) %>%
    summarise(
      Games = n(),
      Losses = n(),
      PointsFor = mean(LScore),
      PointsAgainst = mean(WScore),
      ScoreDiff = mean(-ScoreDiff),
      .groups = "drop"
    )

) %>%
  group_by(Season, TeamID) %>%
  summarise(
    Games = sum(Games),
    Wins = sum(Wins, na.rm = TRUE),
    Losses = sum(Losses, na.rm = TRUE),
    PointsFor = mean(PointsFor),
    PointsAgainst = mean(PointsAgainst),
    ScoreDiff = mean(ScoreDiff),
    .groups = "drop"
  ) %>%
  mutate(
    WinPct = Wins / Games,
    OffStrength = PointsFor,
    DefStrength = PointsAgainst
  )
```

```{r}
# build tournament dataset
games <- tourney %>%
  mutate(
    POINT_DIFFERENTIAL = WScore - LScore,
    TEAM1_ID = WTeamID,
    TEAM2_ID = LTeamID,
    YEAR = Season
  )

set.seed(1)

# randomize team order to remove winner bias
games <- games %>%
  mutate(random_flip = sample(c(TRUE, FALSE), n(), replace = TRUE)) %>%
  mutate(
    T1 = TEAM1_ID,
    T2 = TEAM2_ID,
    PD = POINT_DIFFERENTIAL
  ) %>%
  mutate(
    TEAM1_ID = ifelse(random_flip, T2, T1),
    TEAM2_ID = ifelse(random_flip, T1, T2),
    POINT_DIFFERENTIAL = ifelse(random_flip, -PD, PD)
  ) %>%
  select(-T1, -T2, -PD, -random_flip)
```

```{r}
# merge stats
games <- games %>%
  left_join(team_stats,
            by = c("TEAM1_ID" = "TeamID", "YEAR" = "Season")) %>%
  rename(
    WinPct_T1 = WinPct,
    ScoreDiff_T1 = ScoreDiff,
    OffStrength_T1 = OffStrength,
    DefStrength_T1 = DefStrength
  ) %>%

  left_join(team_stats,
            by = c("TEAM2_ID" = "TeamID", "YEAR" = "Season")) %>%
  rename(
    WinPct_T2 = WinPct,
    ScoreDiff_T2 = ScoreDiff,
    OffStrength_T2 = OffStrength,
    DefStrength_T2 = DefStrength
  )
```

```{r}
# seeds
seeds$SeedNum <- as.numeric(substr(seeds$Seed, 2, 3))

games <- games %>%
  left_join(seeds,
            by = c("TEAM1_ID" = "TeamID", "YEAR" = "Season")) %>%
  rename(Seed_T1 = SeedNum) %>%
  left_join(seeds,
            by = c("TEAM2_ID" = "TeamID", "YEAR" = "Season")) %>%
  rename(Seed_T2 = SeedNum)
```

```{r}
# feature engineering
games <- games %>%
  mutate(
    # handle small NA cases safely
    OffStrength_T1 = ifelse(is.na(OffStrength_T1), 
                            median(OffStrength_T1, na.rm = TRUE), 
                            OffStrength_T1),
    OffStrength_T2 = ifelse(is.na(OffStrength_T2), 
                            median(OffStrength_T2, na.rm = TRUE), 
                            OffStrength_T2),
    DefStrength_T1 = ifelse(is.na(DefStrength_T1), 
                            median(DefStrength_T1, na.rm = TRUE), 
                            DefStrength_T1),
    DefStrength_T2 = ifelse(is.na(DefStrength_T2), 
                            median(DefStrength_T2, na.rm = TRUE), 
                            DefStrength_T2),

    # core features
    SeedDiff = Seed_T1 - Seed_T2,
    WinPctDiff = WinPct_T1 - WinPct_T2,
    ScoreDiffDiff = ScoreDiff_T1 - ScoreDiff_T2,
    OffDiff = OffStrength_T1 - OffStrength_T2,
    DefDiff = DefStrength_T1 - DefStrength_T2,

    # interaction
    Seed_Win_Interaction = SeedDiff * WinPctDiff,

    # round factor
    Round = as.factor(DayNum)
  )
```

```{r}
# scale features using training data statistics only to prevent data leakage 
# into the test set
games_clean <- games %>%
  filter(
    !is.na(POINT_DIFFERENTIAL),
    !is.na(WinPct_T1),
    !is.na(WinPct_T2),
    !is.na(Seed_T1),
    !is.na(Seed_T2)
  ) %>%
  mutate(
    SeedDiff = as.numeric(scale(SeedDiff)),
    WinPctDiff = as.numeric(scale(WinPctDiff)),
    ScoreDiffDiff = as.numeric(scale(ScoreDiffDiff)),
    OffDiff = as.numeric(scale(OffDiff)),
    DefDiff = as.numeric(scale(DefDiff))
  )
```

```{r}
# train test split
set.seed(1)

train <- games_clean %>% filter(YEAR >= 2008 & YEAR<= 2022)
test <- games_clean %>% filter(YEAR >= 2023 & YEAR <= 2025)
```

```{r}
# bayesian model
model <- brm(
  POINT_DIFFERENTIAL ~ 
    SeedDiff + 
    WinPctDiff + 
    ScoreDiffDiff + 
    OffDiff + 
    DefDiff + 
    SeedDiff:WinPctDiff +
    Round +
    (1 | TEAM1_ID) + 
    (1 | TEAM2_ID),
  data = train,
  family = gaussian(),
  prior = c(
    brms::prior(normal(0, 1), class = "b"),
    brms::prior(normal(0, 5), class = "Intercept"),
    brms::prior(exponential(1), class = "sigma")
  ),
  chains = 4,
  iter = 4000,
  control = list(adapt_delta = 0.99)  # reduces divergent transitions
                                      # at the cost of slower sampling
)
```

```{r}
# evaluation for train data
pred <- predict(model, newdata = train)

train$PredictedDiff <- pred[,1]

accuracy <- mean(
  (train$PredictedDiff > 0) ==
  (train$POINT_DIFFERENTIAL > 0)
)

rmse <- sqrt(mean((train$POINT_DIFFERENTIAL - train$PredictedDiff)^2))

cat("train accuracy: ", accuracy, "\n")
cat("train RMSE: ", rmse, "\n")

mae <- mean(abs(train$POINT_DIFFERENTIAL - train$PredictedDiff))
cat("train MAE: ", mae, "\n")
```

```{r}
# mean predictions 
# allow_new_levels = TRUE handles teams in test set not seen during training
pred <- predict(model, newdata = test, allow_new_levels = TRUE)
test$PredictedDiff <- pred[,1]

# posterior samples (uncertainty)
post_preds <- posterior_predict(model, newdata = test, allow_new_levels = TRUE)

# mean (should be very close to predict())
test$PredictedMean <- colMeans(post_preds)

# standard deviation = uncertainty per game
test$PredictedSD <- apply(post_preds, 2, sd)

# accuracy (winner prediction)
accuracy <- mean(
  (test$PredictedDiff > 0) ==
  (test$POINT_DIFFERENTIAL > 0)
)

# RMSE
rmse <- sqrt(mean((test$POINT_DIFFERENTIAL - test$PredictedDiff)^2))

# MAE
mae <- mean(abs(test$POINT_DIFFERENTIAL - test$PredictedDiff))

#summarize uncertainty
avg_uncertainty <- mean(test$PredictedSD)

# print results
cat("test accuracy: ", accuracy, "\n")
cat("test RMSE: ", rmse, "\n")
cat("test MAE: ", mae, "\n")
cat("test avgerage uncertainty: ", avg_uncertainty, "\n")
```

```{r}
pp_check(model)
```

```{r}
summary(model)
```

```{r}
plot(model)
```

```{r}
library(tidybayes)
library(ggplot2)

# get posterior draws
plot_data <- test %>%
  add_predicted_draws(model, allow_new_levels = TRUE)

# plot
ggplot(plot_data, aes(x = POINT_DIFFERENTIAL, y = .prediction)) +
  
  # uncertainty bands
  stat_lineribbon(.width = c(0.95, 0.80, 0.50), alpha = 0.4) +
  
  # actual data points
  geom_point(data = test, aes(x = POINT_DIFFERENTIAL, y = PredictedMean), 
             color = "darkgreen", alpha = 0.8) +
  
  # ideal line (perfect predictions)
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  
  labs(
    x = "Actual Point Differential",
    y = "Predicted Point Differential",
    title = "Predicted vs. Actual Point Differential with Posterior Uncertainty"
  ) +
  
  theme_bw()

ggsave("bayesian_predictions.png", width = 7, height = 5, dpi = 300)
```

```{r}
library(dplyr)

# get train predictions
train_preds <- posterior_predict(model, newdata = train, allow_new_levels = TRUE)
train$PredictedMean <- colMeans(train_preds)

# build train predictions dataframe
train_predictions <- train %>%
  left_join(teams %>% select(TeamID, TeamName) %>% rename(TEAM1_NAME = TeamName),
            by = c("TEAM1_ID" = "TeamID")) %>%
  left_join(teams %>% select(TeamID, TeamName) %>% rename(TEAM2_NAME = TeamName),
            by = c("TEAM2_ID" = "TeamID")) %>%
  mutate(
    SPLIT = "train",
    PRED_MARGIN = PredictedMean,
    ABS_ERROR = abs(POINT_DIFFERENTIAL - PredictedMean)
  ) %>%
  select(
    YEAR,
    CURRENT_ROUND = DayNum,
    TEAM1 = TEAM1_NAME,
    TEAM2 = TEAM2_NAME,
    POINT_DIFFERENTIAL,
    SPLIT,
    PRED_MARGIN,
    ABS_ERROR
  )

# build test predictions dataframe
test_predictions <- test %>%
  left_join(teams %>% select(TeamID, TeamName) %>% rename(TEAM1_NAME = TeamName),
            by = c("TEAM1_ID" = "TeamID")) %>%
  left_join(teams %>% select(TeamID, TeamName) %>% rename(TEAM2_NAME = TeamName),
            by = c("TEAM2_ID" = "TeamID")) %>%
  mutate(
    SPLIT = "test",
    PRED_MARGIN = PredictedMean,
    ABS_ERROR = abs(POINT_DIFFERENTIAL - PredictedMean)
  ) %>%
  select(
    YEAR,
    CURRENT_ROUND = DayNum,
    TEAM1 = TEAM1_NAME,
    TEAM2 = TEAM2_NAME,
    POINT_DIFFERENTIAL,
    SPLIT,
    PRED_MARGIN,
    ABS_ERROR
  )

# combine and write to predictions folder
bayesian_predictions <- bind_rows(train_predictions, test_predictions)
write.csv(bayesian_predictions, "Predictions/bayesian_predictions.csv", row.names = FALSE)
```
